---
title: "predictcultivar package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{predictcultivar package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_chunk$set(show.col_types = FALSE)
set.seed(123)
```

```{r setup}
library(predictcultivar)
```
## Introduction to  `predictcultivar`
Welcome to our package! It was designed to streamline our data analysis project and produce code that can be easily used by others. This package contains 6 functions to take your project from Exploratory Data Analysis to finished, tuned `knn` model.

### Data: 
To demonstrate the package, we will use the dataset `mtcars` in R. 

```{r load-mtcars, echo=FALSE}
# Load the mtcars dataset
data(mtcars)

# Display the first few rows of the dataset
head(mtcars)
```

This dataset contains information about 32 cars, such as their weight (`wt`) and fuel consumption (`mpg`). For this demonstration, we will aim to predict what type of transmission a car has (automatic or manual) based on `wt` and `mpg`. 

### Calculate summary statistics with `calc_stats()`
This function allows you to calculate the mean and standard deviation for all numerical columns in a dataframe.

```{r}
calc_stats(mtcars)
```

### Create boxplots with `create_boxplot()`
This function produces a boxplot based on the data and variables it is given.

```{r}
create_boxplot(mtcars,"wt","mpg")
```

The function takes the variable names as strings.


### Create scatterplots with `create_scatter()`
This function produces a scatterplot of `x` versus `y`, with the points coloured according to the variable in `color_var`.

```{r}
create_scatter(mtcars,"wt","mpg","am")
```

The function takes the variable names as strings.


### Store outputs with `create_output_dir()`
This function can be used for making sure a directory that is passed into a function exists. It creates a directory if it doesn't exist. If the directory already exists, no changes are made.  

```{r}
list.dirs()
```

```{r}
output_dir <-"results/"
create_output_dir(output_dir)
scatterplot <- create_scatter(mtcars,"wt","mpg","am")
ggplot2::ggsave(file.path(output_dir, "scatterplot.png"), scatterplot)
list.dirs()
```

```{r echo=FALSE}
# delete the results directory after running the previous code so that it runs correctly the next time
unlink("results/", recursive = TRUE)
```

### Read and prepare data for models with `read_and_factor()`
Many classification models depend on the response variable being a factor. This function allows you to read data from a csv file and convert variables in the data to factors. 

The function can accept a single variable to convert to a factor, in string form.
```{r echo=FALSE}
write.csv(mtcars,"data/mtcars_data.csv")
```

```{r}
# show the class of `am` before it is factored
print(paste("Class of `am` before read_and_factor(): " , class(mtcars$am)))

# provide a directory to read data from. This data is the same as that in the mtcars dataset.
mtcars_data_directory <- "data/mtcars_data.csv"
# read the data from the directory and factor `am`
single_factored_data <- read_and_factor(mtcars_data_directory, "am")

# show the class of `am` after it is factored
print(paste("Class of `am` after read_and_factor(): " , class(single_factored_data$am)))
```

The function can also accept a list of variables to convert to factors, in string form. 
```{r}
# show the variable classes before being factored
print(paste("Class of `am` before read_and_factor(): " , class(mtcars$am)))
print(paste("Class of `vs` before read_and_factor(): " , class(mtcars$vs)))

factored_list <- read_and_factor(mtcars_data_directory, c("am", "vs"))

# show the variable classes before after factored
print(paste("Class of `am` before read_and_factor(): " , class(factored_list$am)))
print(paste("Class of `vs` before read_and_factor(): " , class(factored_list$vs)))

```

This works even if some of the variables in the list are not present in the dataframe, as they are ignored and a warning is provided.
```{r}
# show the variable classes before being factored
print(paste("Class of `am` before read_and_factor(): " , class(mtcars$am)))
print(paste("Class of `vs` before read_and_factor(): " , class(mtcars$vs)))

factored_list <- read_and_factor(mtcars_data_directory, c("am", "vs", "fake_var"))

# show the variable classes before after factored
print(paste("Class of `am` before read_and_factor(): " , class(factored_list$am)))
print(paste("Class of `vs` before read_and_factor(): " , class(factored_list$vs)))
```

The function can also read the data without converting anything to factors.
```{r}
unfactored_data <- read_and_factor(mtcars_data_directory)
```
  

### Train, tune and fit a knn classification model with `model_workflow()`
This function streamlines the workflow associated with knn classification models. It:
1. Splits data into training and testing sets
2. Scales and centres all predictors
3. Defines a knn classification specification
4. Uses cross validation to find an optimal `k` based on a specified number of validation folds and range of `k` values to try
5. Fits a new knn classification model based on the optimal `k`
6. Creates a confusion matrix
7. Uses `summarize_classes()` to calculate accuracy, recall, precision and f1 score of the model


```{r}
num_folds <- 5
range_neighbors <- seq(1, 20)
response <- "am"
results <- model_workflow(mtcars, num_folds, range_neighbors, response)

# show to confusion matrix
results[[1]]
# show to dataframe of model metrics
results[[2]]

```


### Score a knn model with `summarize_classes()`
This function is designed to calculate model metrics from a confusion matrix. What makes this function unique is that it works for multiclass, non-numeric confusion matrices as well as binary class, numeric matrices. Based on the standard structure of a confusion matrix produced by `caret::confusionMatrix()`, we can calculate an average value across all classes for any metric present in the confusion matrix.

```{r}
library(predictcultivar)
# define a 3 class confusion matrix according to the standard output of caret::confusionMatrix() 
conf_matrix <- matrix(c(0.3678, 0.3690, 0.30000,
                        0.7134, 0.6108, 0.69591,
                        0.4051, 0.3229, 0.31579,
                        0.6802, 0.6581, 0.68000,
                        0.3466, 0.3347, 0.31873,
                        0.1275, 0.1235, 0.09562,
                        0.6772, 0.69591, 0.6802,
                        0.49795, 0.5406, 0.31873,
                        0.1275, 0.1235, 0.09562,
                        0.3147, 0.3825, 0.30279,
                        0.5406, 0.4899, 0.49795),
                      nrow = 3, byrow = TRUE)
rownames(conf_matrix) <- c("Class 1", "Class 2", "Class 3")
colnames(conf_matrix) <- c("Sensitivity", "Specificity", "Pos Pred Value", "Neg Pred Value", "Precision", "Recall", "F1","Prevalence", "Detection Rate", "Detection Prevalence", "Balanced Accuracy")

calc_recall <- summarize_classes(conf_matrix,3,"Recall")
calc_precision <- summarize_classes(conf_matrix,3,"Recall")
calc_detection_rate <- summarize_classes(conf_matrix,3,"Detection Rate")

# put calculations into a data frame
metrics_df <- data.frame(
    Recall = calc_recall,
    Precision = calc_precision,
    Detection_Rate = calc_detection_rate
  )

metrics_df
```








